Descriptive Complexity for Counting Complexity Classes

### 2017-07-27 LMCS/saluja.tex
Pasada por sección 5.1.

Inspired by the connection between $\shp$ and $\sfo$, a hierarchy of subclases of $\sfo$ was introduced in~\cite{SalujaST95} 
by restricting the alternation of quantifiers in Boolean formulae.
Specifically, the \emph{$\sfo$-hierarchy} consists of the 
the classes $\E{i}$ and $\U{i}$ for every $i \geq 0$, where $\E{i}$ (resp., $\U{i}$) is defined as $\sfo$ but restricting the formulae used to be in $\loge{i}$ (resp., $\logu{i}$).
By definition, we have that $\U{0} = \E{0}$. Moreover, it is shown in~\cite{SalujaST95} that:
\[
\E{0} \; \subsetneq \; \E{1} \; \subsetneq \; \U{1} \; \subsetneq \; \E{2} \; \subsetneq \; \U{2} \; = \; \sfo 
\]
In light of the framework introduced in this paper, natural extensions of these classes are obtained by considering 
$\eqso(\loge{i})$ and $\eqso(\logu{i})$ for every $i \geq 0$, which form the \emph{$\eqso(\fo)$-hierarchy}.
Clearly, we have that $\E{i} \subseteq \QE{i}$ and $\U{i} \subseteq \QU{i}$. Indeed, each formula $\varphi(\bar{X}, \bar{x})$ in $\E{i}$ is equivalent to the formula $\sa{\bar X} \sa{\bar x} \varphi(\bar{X}, \bar{x})$ in $\QE{i}$, and likewise for $\U{i}$ and $\QU{i}$.
But what is the exact relationship between these two hierarchies?
To answer this question, we first introduce two normal forms for $\eqso(\LL)$ that helps us to characterize the expressive power of this quantitative logic.
A formula $\alpha$ in $\eqso(\LL)$ is in \emph{$\LL$-prenex normal form ($\LL$-PNF)} 
if $\alpha$ is of the form
$\sa{\bar{X}} \sa{\bar{x}} \varphi(\bar{X}, \bar{x})$,
where $\bar{X}$ and $\bar{x}$ are sequences of zero or more second-order and first-order variables, respectively, and $\varphi(\bar{X}, \bar{x})$ is a formula in $\LL$. Notice that 
a formula $\varphi(\bar{X}, \bar{x})$ in $\sh{\LL}$ is equivalent to the formula $\sa{\bar X} \sa{\bar x} \varphi(\bar{X}, \bar{x})$ in $\LL$-PNF. 
Moreover, a formula $\alpha$ in $\eqso(\LL)$ is in \emph{$\LL$-sum normal form ($\LL$-SNF)} if $\alpha$ is of the form $\Sigma_{i=1}^n \alpha_i$ where each $\alpha_i$ is in $\LL$-PNF. 
\begin{prop}\label{theo-pnf-snf}
Every formula in $\eqso(\LL)$ can be rewritten in $\LL$-SNF.
\end{prop}
\proof
\input{proof-theo-pnf-snf}
\qed
If a formula is in $\LL$-PNF then clearly the formula is in $\LL$-SNF.
Unfortunately, for some $\LL$ there exist formulae in $\eqso(\LL)$  that cannot be rewritten in $\LL$-PNF.
Therefore, to unveil the relationship between the $\sfo$-hierarchy and the $\eqso(\fo)$-hierarchy, we need to understand the boundary between PNF and SNF. We do this in the following theorem. 
\begin{thm}\label{theo-pi1-pnf}
For $i = 0,1$, there exists a formula $\alpha_i$ in $\QE{i}$ that is not equivalent to any formula in $\Sigma_i$-PNF. 
On the other hand, if $\logu{1} \subseteq \LL$ and $\LL$ is closed under conjunction and disjunction, then every formula in $\eqso(\LL)$ can be rewritten in $\LL$-PNF.

From now on, for every first-order tuple $\bar{x}$ (second-order tuple $\bar{X}$) $\length{\bar{x}}$ as the size of $\bar{x}$ ($\length{\bar{X}}$ as the size of $\bar{X}$).
\end{thm}
\proof
\input{proof-theo-pi1-pnf}
\qed

@@

	\node[rectw, right=0.8cm of n5] (n6) {$\U{1}$}       

@@

	\node[rectw, below=0.3cm of n6] (n7) {$\QU{1}$}       

@@

	\node[rectw, right=1.0cm of n6] (n8) {$\E{2}$}       

@@

	\node[rectw, below=0.3cm of n8] (n9) {$\QE{2}$}       

@@

	\node[rectw, right=1.0cm of n8] (n10) {$\U{2}$}       

@@

	\node[rectw, below=0.3cm of n10] (n11) {$\QU{2}$}       

@@

\vspace{1cm}
	\node[rectw, right=1.2cm of n1] (n2) {};
	\node[rectw, above=0.2cm of n2] (n3) {$\E{1}$}
	\node[rectw, below=0.1cm of n2] (n4) {$\QE{0}$}
	\node[rectw, right=1.2cm of n2] (n5) {$\QE{1}$}
	\node[rectw, right=1.3cm of n5] (n6) {\rotatebox{90}{$=$}}       
	\node[rectw, above=0cm of n6] (n13) {$\U{1}$};
	\node[rectw, below=0cm of n6] (n7) {$\QU{1}$};      
	\node[rectw, right=1.8cm of n6] (n8) {\rotatebox{90}{$=$}}       
	\node[rectw, above=0cm of n8] (n14) {$\E{2}$};
	\node[rectw, below=0cm of n8] (n9) {$\QE{2}$};       
	\node[rectw, right=1.8cm of n8] (n10) {\rotatebox{90}{$=$}}       
	\node[rectw, above=0cm of n10] (n15) {$\U{2}$};
	\node[rectw, below=0cm of n10] (n11) {$\QU{2}$};      
\vspace{-0.4cm}

@@

As a consequence of Proposition~\ref{theo-pnf-snf} and Theorem~\ref{theo-pi1-pnf}, we obtain that $\E{i} \subsetneq \QE{i}$ for $i = 0,1$, and that $\sh{\LL} = \eqso(\LL)$ for $\LL$ equal to  $\Pi_1$, $\Sigma_2$ or $\Pi_2$. The following proposition completes our picture of the relationship between the $\sfo$-hierarchy and the $\eqso(\fo)$-hierarchy.
\begin{prop}\label{prop-rest}
The following properties hold:
\begin{itemize}
\item $\QE{0}$ and $\E{1}$ are incomparable, that is, $\E{1} \not\subseteq \QE{0}$ and $\QE{0} \not\subseteq \E{1}$
\item $\QE{1} \subsetneq \QU{1}$
\end{itemize}
\end{prop}
\proof
\input{proof-prop-rest}
\qed
The relationship between the two hierarchies is summarized in Figure \ref{fig-sfo-eqso}.
Our hierarchy and the one proposed in~\cite{SalujaST95} only differ in~$\Sigma_0$ and~$\Sigma_1$. 
Interestingly, we show next that this difference is crucial for finding classes of functions with easy decision versions and good closure properties.

>>>>

Inspired by the connection between $\shp$ and $\sfo$, a hierarchy of subclases of $\sfo$ was introduced in~\cite{SalujaST95} 
by restricting the alternation of quantifiers in Boolean formulae.
Specifically, the \emph{$\sfo$-hierarchy} consists of the 
the classes $\E{i}$ and $\U{i}$ for every $i \geq 0$, where $\E{i}$ (resp., $\U{i}$) is defined as $\sfo$ but restricting the formulae used to be in $\loge{i}$ (resp., $\logu{i}$).
By definition, we have that $\U{0} = \E{0}$. Moreover, it is shown in~\cite{SalujaST95} that:
\[
\E{0} \; \subsetneq \; \E{1} \; \subsetneq \; \U{1} \; \subsetneq \; \E{2} \; \subsetneq \; \U{2} \; = \; \sfo 
\]
In light of the framework introduced in this paper, natural extensions of these classes are obtained by considering 
$\eqso(\loge{i})$ and $\eqso(\logu{i})$ for every $i \geq 0$, which form the \emph{$\eqso(\fo)$-hierarchy}.
Clearly, we have that $\E{i} \subseteq \QE{i}$ and $\U{i} \subseteq \QU{i}$. Indeed, each formula $\varphi(\bar{X}, \bar{x})$ in $\E{i}$ is equivalent to the formula $\sa{\bar X} \sa{\bar x} \varphi(\bar{X}, \bar{x})$ in $\QE{i}$, and likewise for $\U{i}$ and $\QU{i}$.
But what is the exact relationship between these two hierarchies?
To answer this question, we first introduce two normal forms for $\eqso(\LL)$ that helps us to characterize the expressive power of this quantitative logic.
A formula $\alpha$ in $\eqso(\LL)$ is in \emph{$\LL$-prenex normal form ($\LL$-PNF)} 
if $\alpha$ is of the form
$\sa{\bar{X}} \sa{\bar{x}} \varphi(\bar{X}, \bar{x})$,
where $\bar{X}$ and $\bar{x}$ are sequences of zero or more second-order and first-order variables, respectively, and $\varphi(\bar{X}, \bar{x})$ is a formula in $\LL$. Notice that 
a formula $\varphi(\bar{X}, \bar{x})$ in $\sh{\LL}$ is equivalent to the formula $\sa{\bar X} \sa{\bar x} \varphi(\bar{X}, \bar{x})$ in $\LL$-PNF. 
Moreover, a formula $\alpha$ in $\eqso(\LL)$ is in \emph{$\LL$-sum normal form ($\LL$-SNF)} if $\alpha$ is of the form $\Sigma_{i=1}^n \alpha_i$ where each $\alpha_i$ is in $\LL$-PNF. 
\begin{prop}\label{theo-pnf-snf}
Every formula in $\eqso(\LL)$ can be rewritten in $\LL$-SNF.
\end{prop}
\proof
\input{proof-theo-pnf-snf}
\qed
If a formula is in $\LL$-PNF then clearly the formula is in $\LL$-SNF.
Unfortunately, for some $\LL$ there exist formulae in $\eqso(\LL)$  that cannot be rewritten in $\LL$-PNF.
Therefore, to unveil the relationship between the $\sfo$-hierarchy and the $\eqso(\fo)$-hierarchy, we need to understand the boundary between PNF and SNF. We do this in the following theorem. 
\begin{thm}\label{theo-pi1-pnf}
For $i = 0,1$, there exists a formula $\alpha_i$ in $\QE{i}$ that is not equivalent to any formula in $\Sigma_i$-PNF. 
On the other hand, if $\logu{1} \subseteq \LL$ and $\LL$ is closed under conjunction and disjunction, then every formula in $\eqso(\LL)$ can be rewritten in $\LL$-PNF. 
\end{thm}
\proof
\input{proof-theo-pi1-pnf}
\qed

@@

	\node[rectw, right=0.8cm of n5] (n6) {$\QU{1}$}       

@@

	\node[rectw, below=0.3cm of n6] (n7) {$\U{1}$}       

@@

	\node[rectw, right=1.0cm of n6] (n8) {$\QE{2}$}       

@@

	\node[rectw, below=0.3cm of n8] (n9) {$\E{2}$}       

@@

	\node[rectw, right=1.0cm of n8] (n10) {$\QU{2}$}       

@@

	\node[rectw, below=0.3cm of n10] (n11) {$\U{2}$}       

@@

%\vspace{1cm}
%	\begin{tikzpicture}
%	\node[rectw] (n1) {$\E{0}$};
%	\node[rectw, right=1.2cm of n1] (n2) {};
%	\node[rectw, above=0.2cm of n2] (n3) {$\E{1}$}
%	\node[rectw, below=0.1cm of n2] (n4) {$\QE{0}$}
%	\node[rectw, right=1.2cm of n2] (n5) {$\QE{1}$}
%	\node[rectw, right=1.3cm of n5] (n6) {\rotatebox{90}{$=$}}       
%	\node[rectw, above=0cm of n6] (n13) {$\U{1}$};
%	\node[rectw, below=0cm of n6] (n7) {$\QU{1}$};      
%	\node[rectw, right=1.8cm of n6] (n8) {\rotatebox{90}{$=$}}       
%	edge[draw=white] node {$\subsetneq$} (n6);
%	\node[rectw, above=0cm of n8] (n14) {$\E{2}$};
%	\node[rectw, below=0cm of n8] (n9) {$\QE{2}$};       
%	\node[rectw, right=1.8cm of n8] (n10) {\rotatebox{90}{$=$}}       
%	edge[draw=white] node {$\subsetneq$} (n8);
%	\node[rectw, above=0cm of n10] (n15) {$\U{2}$};
%	\node[rectw, below=0cm of n10] (n11) {$\QU{2}$};      
%	edge[draw=white] node {\rotatebox{90}{$=$}} (n10); 
%	\node[rectw, right=0.5cm of n10] (n12) {$\sfo$}       
%	\end{tikzpicture}

@@

As a consequence of Proposition~\ref{theo-pnf-snf} and Theorem~\ref{theo-pi1-pnf}, we obtain that $\E{i} \subsetneq \QE{i}$ for $i = 0,1$, and that $\sh{\LL} = \eqso(\LL)$ for $\LL$ equal to  $\Pi_1$, $\Sigma_2$ or $\Pi_2$. The following proposition completes our picture of the relationship between the $\sfo$-hierarchy and the $\eqso(\fo)$-hierarchy.
\begin{prop}\label{prop-rest}
The following properties hold:
\begin{itemize}
\item $\QE{0}$ and $\E{1}$ are incomparable, that is, $\E{1} \not\subseteq \QE{0}$ and $\QE{0} \not\subseteq \E{1}$,
\item $\QE{1} \subsetneq \QU{1}$.
\end{itemize}
\end{prop}
\proof
\input{proof-prop-rest}
\qed
The relationship between the two hierarchies is summarized in Figure \ref{fig-sfo-eqso}.
Our hierarchy and the one proposed in~\cite{SalujaST95} only differ in~$\Sigma_0$ and~$\Sigma_1$. 
Interestingly, we show next that this difference is crucial for finding classes of functions with easy decision versions and good closure properties.
